Breast Cancer Diagnosis Classification Analysis

### 2022-12-15 paper.rmd

*Insert text here*

@@

*Insert text here regarding the GGPairs Plot*

@@

knn_spec <- the_model <- nearest_neighbor(neighbors = tune("neighbors")) %>%
  set_engine("kknn") %>% 
  set_mode("classification")

@@

*Insert text here*

@@

*Insert text here*

>>>>

After removing the `ID` column, we were left with `31 variables`. 
For this project our response variable is `diagnosis` as this variable tells whether an observation is Malignant (cancerous cells): `M` or Benign (noncancerous cells): `B`.
The predictor variables make up the remaining 30 columns; however there are only actually 10 variables that each have 3 measurements.
These measurements are `mean`, `standard error (SE)`, and `worst` (the mean of the 3 largest values of a variables).

Below is a brief explanation of what each of the 10 variables represent:

- Radius: Average Distance from cell center to cell perimeter
- Texture: Standard deviation of gray-scale values; brightness of pixel of cell
- Perimeter: Distance around nucleus boundary
- Area: Area of the nucleus
- Smoothness: Variation in cell's radial lengths
- Compactness: The Perimeter^2/Area
- Concavity: Size of the indention in nucleus boundary
- Concave Points: Number of points on indented section of nucleus boundary
- Symmetry: Deviation of the nuclei shape from the ideal measurement
- Fractal Dimension: Measurement of irregularity in nucleus boundary

@@

Another important thing to note about the data is the variables included. There are three different measurements for the ten predictor variables: mean value, standard error, and "worst" value (mean of 3 largest measurements). As shown the correlation plot below, these three measurements are all correlated to each other, but still hold a strong relationship with the diagnosis response variable. The relation between the mean value and "worst" value is almost completely correlated which makes sense as both are averages however, the "worst" value only uses the three largest measurements. This is important to know when trying to interpret feature effect, importance and selection.

@@

knn_spec <- nearest_neighbor(neighbors = tune("neighbors")) %>%
  set_engine("kknn") %>% 
  set_mode("classification")

@@

```{r, echo=FALSE, message=FALSE, warning=FALSE, results=FALSE}
# Import packages
library(tidyverse)
library(ggplot2)
library(GGally)
library(knitr)
library(boot)
library(caret)
library(randomForest)
library(e1071)
library(class)
#
library(dplyr)
library(leaps)
library(caTools)
library(tidymodels)
library(glmnet)
library(forecast)
library(plyr)
library(readr)
library(repr)
library(kableExtra)

dataTemp <- read.csv("breast_cancer_diagnostic_data.csv")
datatemp2 <- dataTemp %>% select_if(~ !any(is.na(.)))
dataTot <- datatemp2[,!names(datatemp2) %in% c("id")]

testtemp <- read.csv("test.csv")
testTot <- testtemp[,!names(testtemp) %in% c("id")]

traintemp <- read.csv("training.csv")
trainTot <- traintemp[,!names(traintemp) %in% c("id")]
```

While Austin and Danny did Classification Analysis on `Type-II error rate`, I decided to do a Regression Analysis using the variable importance from their analysis's to compute the `training` and `test` error rate of those variables. 
This is to see how much error exists in a variable that is important in calculating if a cell is Malignant or Benign.
However, I will be focused on `test` error in my results as its more important to see how our results will be on future data rather than what we have already trained. The regression models that I will be using are:

    - Linear Regression
    - Ridge Regression
    - Lasso Regression

## Variables

In order to find the variables to run the Regression Analysis on, I had Austin and Danny run a variable importance plot on the data they collected and took the top 5 variables from each of their results, which are listed below.

- Austin Top 5 Importance Variables:

  1. *`area_mean`
  2. *`area_worst`
  3. *`concave.points_mean`
  4. `texture_mean`
  5. *`radius_worst`

- Danny Top 5 Importance Variables

  1. `concave.points_worst`
  2. *`concave.points_mean`
  3. *`area_worst`
  4. *`area_mean`
  5. *`radius_worst`

After I got both Austin's and Danny's top 5 importance variables, I noticed that there were 4 common variables in each of their results (designated with a '*' in the results above).
I then decided to see how the 4 common importance variables correlate to one another which can be seen in the pairwise scatterplot below:

```{r correlation_plot, results=FALSE, echo=FALSE, message=FALSE, warning=FALSE}
corr_df <- dataTot[,c( 'area_worst', 'area_mean', 'concave.points_mean', 'radius_worst')]
ggpairs(corr_df) + labs(title = "Figure 1: Pairwise Scatterplot of Common Importance Variables")
```

What I found interesting was that `area_worst` & `raduis_worst` were more correlated that `area_worst` & `area_mean`. What I initially thought before making `Figure 1` was that `area_worst` & `area_mean` would be more correlated than the other 2 variables as they are both under the same variable category.

## area_mean

```{r, results=FALSE, echo=FALSE, message=FALSE, warning=FALSE}
#Train
am_lm <- lm(area_mean ~ ., data = trainTot)
pred_amtrain<- predict(am_lm, trainTot)
amtrain_log <- postResample(pred_amtrain, trainTot$area_mean)
#amtrain_log
```
```{r, results=FALSE, echo=FALSE, message=FALSE, warning=FALSE}
#Test
pred_amtest <- predict(am_lm, testTot)
amtest_log <- postResample(pred_amtest, testTot$area_mean)
#amtest_log
#summary(am_lm)
```

```{r, results=FALSE, echo=FALSE, message=FALSE, warning=FALSE}
dummy <- dummyVars(area_mean~., data = dataTot[,-1])
x = as.matrix(predict(dummy, newdata = trainTot[,-1]))
y_train = trainTot$area_mean

x_test = as.matrix(predict(dummy, newdata = testTot[,-1]))
y_test = testTot$area_mean

lambdas <- 10^seq(2, -3, by = -.1)
ridge_reg = glmnet(x, y_train, alpha = 0, family = 'gaussian', lambda = lambdas)
cv_ridge <- cv.glmnet(x, y_train, alpha = 0, lambda = lambdas)
optimal_lambda <- cv_ridge$lambda.min

eval_results <- function(true, predicted, df) {
  SSE <- sum((predicted - true)^2)
  SST <- sum((true - mean(true))^2)
  R_square <- 1 - SSE / SST
  RMSE = sqrt(SSE/nrow(df))

  # Model performance metrics
data.frame(
  RMSE = RMSE,
  Rsquared = R_square)

}

# Prediction and evaluation on train data
predictions_train <- predict(ridge_reg, s = optimal_lambda, newx = x)
amr_train <- eval_results(y_train, predictions_train, trainTot)

# Prediction and evaluation on test data
predictions_test <- predict(ridge_reg, s = optimal_lambda, newx = x_test)
amr_test <- eval_results(y_test, predictions_test, testTot)

#print(amr_train)
#print(amr_test)
```

```{r, results=FALSE, echo=FALSE, message=FALSE, warning=FALSE}
lambdas <- 10^seq(2, -3, by = -.1)
lasso_reg <- cv.glmnet(x, y_train, alpha = 1, lambda = lambdas, standardize = TRUE, nfolds = 10)

lambda_best <- lasso_reg$lambda.min 
lasso_model <- glmnet(x, y_train, alpha = 1, lambda = lambda_best, standardize = TRUE)

predictions_train <- predict(lasso_model, s = lambda_best, newx = x)
aml_train <- eval_results(y_train, predictions_train, trainTot)

predictions_test <- predict(lasso_model, s = lambda_best, newx = x_test)
aml_test <- eval_results(y_test, predictions_test, testTot)

#print(aml_train)
#print(aml_test)
```

For `area_mean` after running the 3 regression models, I found the lowest `test error` is from the ridge regression model with an `RMSE` of `23.70795` with a `Rsquared` of `99.489%`. 
This mean that using ridge regression, we can explain `99.489%` of the variance in `area_mean` in Austin's and Danny's models.
The ridge regression corresponding `training error` is `17.87197` with a `Rsquared` of 	`99.759%`.
This corresponds to that `99.759%` of our variance in `area_mean` was trained in Austin's and Danny's models.

```{r area_mean_table, echo=FALSE, message=FALSE, warning=FALSE}
table_am <- as_tibble(rbind(amtrain_log,
      amtest_log,
      amr_train,
      amr_test,
      aml_train,
      aml_test)) %>%
    mutate(model = c('Linear Train Error',  'Linear Test Error','Ridge Train Error', 'Ridge Test Error', 'Lasso Train Error', 'Lasso Test Error')) %>%
    select(model, RMSE, Rsquared)
table_am %>% 
  kbl(digits = 5 ) %>% 
  kable_styling(bootstrap_options = "striped") %>%
  kable_styling(latex_options = "HOLD_position")
```

## Area_worst

```{r, echo=FALSE, message=FALSE, warning=FALSE, results=FALSE}
#Test
aw_lm <- lm(area_worst ~ ., data = trainTot)
pred_awtest <- predict(aw_lm, testTot)
awtest_log <- postResample(pred_awtest, testTot$area_worst)
#awtest_log
#summary(am_lm)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE, results=FALSE}
#Train
pred_awtrain<- predict(aw_lm, trainTot)
awtrain_log <- postResample(pred_awtrain, trainTot$area_worst)
#awtrain_log
```

```{r, echo=FALSE, message=FALSE, warning=FALSE, results=FALSE}
dummy <- dummyVars(area_worst~., data = dataTot[,-1])
x = as.matrix(predict(dummy, newdata = trainTot[,-1]))
y_train = trainTot$area_worst

x_test = as.matrix(predict(dummy, newdata = testTot[,-1]))
y_test = testTot$area_worst

lambdas <- 10^seq(2, -3, by = -.1)
ridge_reg = glmnet(x, y_train, alpha = 0, family = 'gaussian', lambda = lambdas)
cv_ridge <- cv.glmnet(x, y_train, alpha = 0, lambda = lambdas)
optimal_lambda <- cv_ridge$lambda.min

eval_results <- function(true, predicted, df) {
  SSE <- sum((predicted - true)^2)
  SST <- sum((true - mean(true))^2)
  R_square <- 1 - SSE / SST
  RMSE = sqrt(SSE/nrow(df))

  # Model performance metrics
data.frame(
  RMSE = RMSE,
  Rsquared = R_square)

}

# Prediction and evaluation on train data
predictions_train <- predict(ridge_reg, s = optimal_lambda, newx = x)
awr_train <- eval_results(y_train, predictions_train, trainTot)

# Prediction and evaluation on test data
predictions_test <- predict(ridge_reg, s = optimal_lambda, newx = x_test)
awr_test <- eval_results(y_test, predictions_test, testTot)

#print(awr_train)
#print(awr_test)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE, results=FALSE}
lambdas <- 10^seq(2, -3, by = -.1)
lasso_reg <- cv.glmnet(x, y_train, alpha = 1, lambda = lambdas, standardize = TRUE, nfolds = 10)

lambda_best <- lasso_reg$lambda.min 
lasso_model <- glmnet(x, y_train, alpha = 1, lambda = lambda_best, standardize = TRUE)

predictions_train <- predict(lasso_model, s = lambda_best, newx = x)
awl_train <- eval_results(y_train, predictions_train, trainTot)

predictions_test <- predict(lasso_model, s = lambda_best, newx = x_test)
awl_test <- eval_results(y_test, predictions_test, testTot)

#print(awl_train)
#print(awl_test)
```

In `area_worst` after running the 3 regression models, I found the lowest `test error` is from the ridge regression model with an `RMSE` of `42.28132` with a `Rsquared` of `99.288%`. 
This mean that using ridge regression, we can explain `99.288%` of the variance in `area_worst` in Austin's and Danny's models.
The ridge regression corresponding `training error` is `27.36780` with a `Rsquared` of `99.799%`.
This corresponds to that `99.799%` of our variance in `area_worst` was trained in Austin's and Danny's models.

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.pos = "H", out.extra = ""}
table_aw <- as_tibble(rbind(awtrain_log,
      awtest_log,
      awr_train,
      awr_test,
      awl_train,
      awl_test)) %>%
    mutate(model = c('Linear Train Error',  'Linear Test Error','Ridge Train Error', 'Ridge Test Error', 'Lasso Train Error', 'Lasso Test Error')) %>%
    select(model, RMSE, Rsquared)
table_aw %>% 
  kbl(digits = 5 ) %>% 
  kable_styling(bootstrap_options = "striped") %>%
  kable_styling(latex_options = "HOLD_position")
```

## concave.points_mean

```{r, echo=FALSE, message=FALSE, warning=FALSE, results=FALSE}
#Train
cvptsm_lm <- lm(concave.points_mean ~ ., data = trainTot)
pred_cvptsm_train<- predict(cvptsm_lm, trainTot)
cvptsm_train_log <- postResample(pred_cvptsm_train, trainTot$concave.points_mean)
#cvptsm_train_log
```

```{r, echo=FALSE, message=FALSE, warning=FALSE, results=FALSE}
#Test
pred_cvptsm_test <- predict(cvptsm_lm, testTot)
cvptsm_test_log <- postResample(pred_cvptsm_test, testTot$concave.points_mean)
#cvptsm_test_log
#summary(am_lm)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE, results=FALSE, fig.pos = "H", out.extra = ""}
dummy <- dummyVars(concave.points_mean~., data = dataTot[,-1])
x = as.matrix(predict(dummy, newdata = trainTot[,-1]))
y_train = trainTot$concave.points_mean

x_test = as.matrix(predict(dummy, newdata = testTot[,-1]))
y_test = testTot$concave.points_mean

lambdas <- 10^seq(2, -3, by = -.1)
ridge_reg = glmnet(x, y_train, alpha = 0, family = 'gaussian', lambda = lambdas)
cv_ridge <- cv.glmnet(x, y_train, alpha = 0, lambda = lambdas)
optimal_lambda <- cv_ridge$lambda.min

eval_results <- function(true, predicted, df) {
  SSE <- sum((predicted - true)^2)
  SST <- sum((true - mean(true))^2)
  R_square <- 1 - SSE / SST
  RMSE = sqrt(SSE/nrow(df))

  # Model performance metrics
data.frame(
  RMSE = RMSE,
  Rsquared = R_square)

}

# Prediction and evaluation on train data
predictions_train <- predict(ridge_reg, s = optimal_lambda, newx = x)
cvptsmr_train <- eval_results(y_train, predictions_train, trainTot)

# Prediction and evaluation on test data
predictions_test <- predict(ridge_reg, s = optimal_lambda, newx = x_test)
cvptsmr_test <- eval_results(y_test, predictions_test, testTot)

#print(cvptsmr_train)
#print(cvptsmr_test)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE, results=FALSE}
lambdas <- 10^seq(2, -3, by = -.1)
lasso_reg <- cv.glmnet(x, y_train, alpha = 1, lambda = lambdas, standardize = TRUE, nfolds = 10)

lambda_best <- lasso_reg$lambda.min 
lasso_model <- glmnet(x, y_train, alpha = 1, lambda = lambda_best, standardize = TRUE)

predictions_train <- predict(lasso_model, s = lambda_best, newx = x)
cvptsml_train <- eval_results(y_train, predictions_train, trainTot)

predictions_test <- predict(lasso_model, s = lambda_best, newx = x_test)
cvptsml_test <- eval_results(y_test, predictions_test, testTot)

#print(cvptsml_train)
#print(cvptsml_test)
```

With `concave.points_mean` after running the 3 regression models, I found the lowest `test error` is from the linear regression model with an `RMSE` of `0.00574` with a `Rsquared` of `97.847%`. 
This mean that using ridge regression, we can explain `97.847%` of the variance in `concave.points_mean` in Austin's and Danny's models.
The ridge regression corresponding `training error` is `0.00467` with a `Rsquared` of `98.532%`.
This corresponds to that `98.532%` of our variance in `concave.points_mean` was trained in Austin's and Danny's models.

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.pos = "H", out.extra = ""}
table_am <- as_tibble(rbind(cvptsm_train_log,
      cvptsm_test_log,
      cvptsmr_train,
      cvptsmr_test,
      cvptsml_train,
      cvptsml_test)) %>%
    mutate(model = c('Linear Train Error',  'Linear Test Error','Ridge Train Error', 'Ridge Test Error', 'Lasso Train Error', 'Lasso Test Error')) %>%
    select(model, RMSE, Rsquared)
table_am %>% 
  kbl(digits = 5 ) %>% 
  kable_styling(bootstrap_options = "striped") %>%
  kable_styling(latex_options = "HOLD_position")
```

## radius_worst

```{r, echo=FALSE, message=FALSE, warning=FALSE, results=FALSE}
#Train
rw_lm <- lm(radius_worst ~ ., data = trainTot)
pred_rw_train<- predict(rw_lm, trainTot)
rw_train_log <- postResample(pred_rw_train, trainTot$radius_worst)
#rw_train_log
```
```{r, echo=FALSE, message=FALSE, warning=FALSE, results=FALSE}
#Test
pred_rw_test <- predict(rw_lm, testTot)
rw_test_log <- postResample(pred_rw_test, testTot$radius_worst)
#rw_test_log
#summary(am_lm)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE, results=FALSE, fig.pos = "H", out.extra = ""}
dummy <- dummyVars(radius_worst~., data = dataTot[,-1])
x = as.matrix(predict(dummy, newdata = trainTot[,-1]))
y_train = trainTot$radius_worst

x_test = as.matrix(predict(dummy, newdata = testTot[,-1]))
y_test = testTot$radius_worst

lambdas <- 10^seq(2, -3, by = -.1)
ridge_reg = glmnet(x, y_train, alpha = 0, family = 'gaussian', lambda = lambdas)
cv_ridge <- cv.glmnet(x, y_train, alpha = 0, lambda = lambdas)
optimal_lambda <- cv_ridge$lambda.min

eval_results <- function(true, predicted, df) {
  SSE <- sum((predicted - true)^2)
  SST <- sum((true - mean(true))^2)
  R_square <- 1 - SSE / SST
  RMSE = sqrt(SSE/nrow(df))

  # Model performance metrics
data.frame(
  RMSE = RMSE,
  Rsquared = R_square)

}

# Prediction and evaluation on train data
predictions_train <- predict(ridge_reg, s = optimal_lambda, newx = x)
rwr_train <- eval_results(y_train, predictions_train, trainTot)

# Prediction and evaluation on test data
predictions_test <- predict(ridge_reg, s = optimal_lambda, newx = x_test)
rwr_test <- eval_results(y_test, predictions_test, testTot)

#print(rwr_train)
#print(rwr_test)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE, results=FALSE}
lambdas <- 10^seq(2, -3, by = -.1)
lasso_reg <- cv.glmnet(x, y_train, alpha = 1, lambda = lambdas, standardize = TRUE, nfolds = 10)

lambda_best <- lasso_reg$lambda.min 
lasso_model <- glmnet(x, y_train, alpha = 1, lambda = lambda_best, standardize = TRUE)

predictions_train <- predict(lasso_model, s = lambda_best, newx = x)
rwl_train <- eval_results(y_train, predictions_train, trainTot)

predictions_test <- predict(lasso_model, s = lambda_best, newx = x_test)
rwl_test <- eval_results(y_test, predictions_test, testTot)

#print(rwl_train)
#print(rwl_test)
```

For `radius_worst` after running the 3 regression models, I found the lowest `test error` is from the lasso regression model with an `RMSE` of `0.24120	` with a `Rsquared` of `99.713%`. 
This mean that using ridge regression, we can explain `99.713%` of the variance in `radius_worst` in Austin's and Danny's models.
The ridge regression corresponding `training error` is `0.15907` with a `Rsquared` of `99.900%`.
This corresponds to that `99.900%` of our variance in `radius_worst` was trained in Austin's and Danny's models.

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.pos = "H", out.extra = ""}
table_am <- as_tibble(rbind(rw_train_log,
      rw_test_log,
      rwr_train,
      rwr_test,
      rwl_train,
      rwl_test)) %>%
    mutate(model = c('Linear Train Error',  'Linear Test Error','Ridge Train Error', 'Ridge Test Error', 'Lasso Train Error', 'Lasso Test Error')) %>%
    select(model, RMSE, Rsquared)
table_am %>% 
  kbl(digits = 5 ) %>% 
  kable_styling(bootstrap_options = "striped") %>%
  kable_styling(latex_options = "HOLD_position")
```

@@

From the results of our classification analysis, we are confident not only in our predictions of tumor diagnosis, but reducing the amount of false-negative diagnoses. Through our methods of tuning specifically for Type II error, and lowering the prediction cutoff for malignant tumors, we have shown two successful ways of battling classification data with a preffered error class. This is extremely important in cancer diagnosis, as well as other analyses, as we do not want to tell someone they don't have cancer when they do, as that could delay treatment and further risk their life.

In terms of feature importance and feature selection, there were many limitations in interpretations due to the correlation of the measurement types. We did find that there were a number of the same features outlined by different models, so we can conclude that the variables, regardless of the measurement type, hold importance in classifying a tumor diagnosis.
